---
title: "PCWAG analysis with MOBSTER"
author: "Giulio Caravagna"
date: "11/03/2020"
output: 
  html_document:
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
```


Table of fits run from the data released with the PCAWG (*Pan-cancer Analysis of Whole-Genomes*) cohort. These fits have been first discussed in the paper G.Caravagna et al. *Model-based tumour subclonal deconvolution.*; to appear in Nature Genetics, 2020; see the [biorXiv preprint](https://www.biorxiv.org/content/10.1101/586560v1).

Table files:

* [ICL analysis](PCWAG_Summary_table.csv);
* [reICL analysis](PCWAG_Summary_table_reICL_reduced.csv).

Tools and data used for this analyis:

* [MOBSTER](https://caravagn.github.io/mobster), a model-based method for tumour subclonal deconvolution;
* [BMix](https://caravagn.github.io/BMix), a Binomial mixture model for read counts clustering;
* Data from the [PCAWG cohort](https://docs.icgc.org/pcawg/).

Note that:

* the tables contains the sample id (`samplename`) as of PCAWG convention, `fit` (ICL or reICL) refers to the two distinct analysis discussed in the main paper;
* `karyotype` identifies the segments copy state of the mutations analysed; the CNA calls  are provided by PCAWG, and  `coverage`  is the average depth of the mutations in the karyotype;
* every column starting with `PCAWG` or `BMix` refers to information generated from PCWAG or BMix analysis. 
* all other columns refer to MOBSTER. Columns named `Mean`/ `Variance` refer to parameters of the corresponding mixture components; columns named `N` to the number of mutations in the final output; columns `pi`  to the relative proportion of mutations and columns `K`  are the number of components fit for each statistical model (`K_B` and `K_BB` in BMix are Binomial/ Beta-Binomial mixtures, with `K_Beta` the Beta components fit by MOBSTER); tail information is reported in `tail` columns.



```{r, include=FALSE}
tab = read_csv('PCWAG_Summary_table.csv') %>% select(-starts_with('rcc'), -X1, -X1_1) %>%
  select(-purity_conf_mad, -sample) %>%
  mutate(fit = 'ICL') %>%
  rename(
    PCAWG_purity = purity, 
    PCAWG_ploidy = ploidy, 
    PCAWG_wgd_status = wgd_status, 
    PCAWG_wgd_uncertain = wgd_uncertain
  )

tab = tab[, order(colnames(tab))] %>%
  select(samplename, karyotype, fit, everything()) 
  
# colnames(tab)[c(7:16, 21:31)] = paste0('MOBSTER_', colnames(tab)[c(7:16, 21:31)])
colnames(tab)[c(5, 6)] = paste0('BMix_', colnames(tab)[c(5, 6)])

tab = tab[, order(colnames(tab))] %>%
  select(samplename, karyotype, fit, coverage, everything()) 

tab2 = read_csv('PCWAG_Summary_table_reICL_reduced.csv') %>% select(-starts_with('rcc')) %>%
  mutate(fit = 'reICL') 

tab2 = tab2[, order(colnames(tab2))] %>%
  select(samplename, karyotype, fit, everything()) 
  
# colnames(tab2)[c(4:25)] = paste0('MOBSTER_', colnames(tab2)[c(4:25)])

tab2 = tab2[, order(colnames(tab2))] %>%
  select(samplename, karyotype, fit, everything()) 
```

### First analysis (MOBSTER + BMix)

This is run with `ICL` for model selection.

```{r, echo=FALSE, message=F}
library(DT)
datatable(tab, class = 'cell-border nowrap', filter = 'top',  extensions = 'Buttons', 
          options = list(
            pageLength = 12, 
            AutoWidth = T,
            dom = 'Bfrtip',
            buttons = c('csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE,
            fixedColumns = TRUE
            )
          )
```

### Second analysis (MOBSTER)

This is run with `reICL` for model selection.

```{r, echo=FALSE, message=F, eval=T}
library(DT)
datatable(tab2, class = 'cell-border nowrap', filter = 'top',  extensions = 'Buttons', 
          options = list(
            pageLength = 12, 
            AutoWidth = T,
            dom = 'Bfrtip',
            buttons = c('csv', 'excel', 'pdf', 'print'),
            scrollX = TRUE,
            fixedColumns = TRUE
            )
          )
```


### Plots

We create some of the plots released in the paper, plus some other example visualisations of the results of one of the two analysis. 

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

# We use results from PCWAG_Summary_table.csv
results = readr::read_csv(file = 'PCWAG_Summary_table.csv', col_types = readr::cols())

# Reshape the data to have some nicer names etc
results = results %>% mutate(tail = ifelse(tail, "With tail", "Without tail"))
results$tail = factor(results$tail, levels = c("With tail", "Without tail"))

# Cohort statistics (purity, coverage and burden)
MEDIAN_PURITY = median(results$purity)
MEDIAN_COVERAGE = median(results$coverage)

MB = quantile(results$N, .8)
```

Mutational burden of tail mutations, split by type of fit.

```{r, fig.width = 3, fig.height=4, warning=FALSE, message=FALSE}
results %>%
  filter(N < MB) %>%
  group_by(tail) %>%
  mutate(xn = round(purity, 3),
         yn = round(N / max(N), 3)) %>%
  distinct(xn, yn, .keep_all = TRUE) %>%
  ungroup %>%
  ggplot(aes(y = yn, x = xn, color = tail)) +
  stat_density2d(
    alpha = .8,
    size = .5,
    color = 'black') +
  geom_point(size = 1, alpha = .6) +
  facet_wrap( ~ tail, nrow = 2) +
  mobster:::my_ggplot_theme() +
  guides(color = FALSE, fill = FALSE) +
  labs(
    x = 'Purity (PCWAG)',
    caption = paste0('MB normalised to maximum, n = ', max(results$N), '.'),
    y = paste0('Mutation burden (MB, %)')
  ) +
  scale_color_manual(values = c(`With tail` = 'indianred3', `Without tail` = 'steelblue')) 
```

We can plot the scatterplot that appears in the main text of the paper.

```{r, fig.width = 6, fig.height=4.5, warning=FALSE, message=FALSE}
scatter_cohort = ggplot(results,
       aes(
         x = purity,
         y = coverage,
         color = tail,
         size = pi_Tail
       )) +
  geom_point(alpha = .7) +
  stat_density2d(
    data = results,
    aes(x = purity, y = coverage, linetype = tail),
    alpha = .8,
    size = .3,
    color = 'black'
  ) +
  ylim(0, 150) +
  mobster:::my_ggplot_theme() +
  guides(
    color = guide_legend("", nrow = 2),
    size = guide_legend("Tail", keyheight = 0),
    linetype = FALSE
  ) +
  geom_vline(
    xintercept = MEDIAN_PURITY,
    linetype = 'longdash',
    size = .15,
    color = 'black'
  ) +
  geom_hline(
    yintercept = MEDIAN_COVERAGE,
    linetype = 'longdash',
    size = .15,
    color = 'black'
  ) +
  labs(
    x = 'Purity (PCWAG)',
    y = 'Median coverage',
    title = "Proportion of tail mutations",
    caption = "Showing only samples with median coverage below 150x.",
    subtitle = paste0(
      'PCWAG (n =',
      unique(results$samplename) %>% length(),
      ' samples, ',
      nrow(results),
      ' karyotypes) with ICL'
    )
  ) +
  xlim(0, 1) +
  scale_color_manual(values = c(`With tail` = 'indianred3', `Without tail` = 'steelblue')) +
  geom_label(
    data = data.frame(
      x = c(0.1, .9, .9, 0.1),
      y = c(150, 150, 0, 0),
      label = c("Q1", "Q2", "Q3", "Q4")
    ),
    aes(x = x, y = y, label = label),
    size = 2,
    color = 'white',
    fill = 'black',
    inherit.aes = F
  ) +
  coord_cartesian(clip = 'off')

scatter_cohort %>% print
```

We can split `scatter_cohort` to show the trend by karyotype, with points showing higher coverage for increased copy states.

```{r, fig.width = 10, fig.height=4, warning=FALSE, message=FALSE}
scatter_cohort + facet_wrap(~karyotype, nrow = 1)
```


We can create a gird of discretised statistics, binning X and Y axis. We can computes for every bin minimum and maximum number of cases with a tail, or other summary statistics.
```{r, warning=FALSE, message=FALSE}
SPAN_X = 0.05
SPAN_Y = 25

# Per-bin statistics
EG = expand.grid(x = seq(0, 1, SPAN_X),
                 y = seq(1, 200, SPAN_Y)) %>%
  apply(MARGIN = 1,
        function(p) {
          results %>% filter(purity >= p['x'],
                             purity < p['x'] + SPAN_X,
                             coverage >= p['y'],
                             coverage < p['y'] + SPAN_Y) %>%
            summarise(
              x = p['x'],
              y = p['y'],
              N = n(),
              mean_prop_Tail = mean(pi_Tail, na.rm = T),
              median_prop_Tail = mean(pi_Tail, na.rm = T),
              min_prop_Tail = min(pi_Tail, na.rm = T),
              max_prop_Tail = max(pi_Tail, na.rm = T),
              detected_Tail = sum(pi_Tail > 0, na.rm = T),
              fequency_Tail = detected_Tail / n(),
              mean_K = mean(K_beta, na.rm = T),
              fequency_Subclone = sum(K_beta > 1, na.rm = T) / n()
            )
        })

EG = Reduce(bind_rows, EG)
```

Plot a table of results from the example statistics.

```{r, fig.width = 6.5, fig.height=3.5, warning=FALSE, message=FALSE}
tb = table(results$karyotype)

mycols <- c(
  `1:0` = 'darkblue',
  `1:1` = "forestgreen",
  `2:0` = "#EFC000FF",
  `2:1` = "#CD534CFF",
  `2:2` = 'purple'
)

ggplot() +
  mobster:::my_ggplot_theme() +
  guides(fill = guide_colorbar("Observed probability \n(m, cases with tail) ", barwidth = 6)) +
  labs(
    x = 'Purity (discretisation 5%)', 
    y = 'Median coverage (discretisation 25x)', 
    caption = "Showing only tiles with at least n = 5 observations."
    ) +
  geom_tile(
    data = EG %>%
      mutate(
        detected_Tail = detected_Tail / max(detected_Tail),
        detected_Tail = ifelse(N > 5, detected_Tail, NA),
        fequency_Tail = ifelse(N > 5, fequency_Tail, NA)
      ),
    aes(
      x,
      y,
      fill = fequency_Tail,
      width = 0.04,
      height = 20
    )
  ) +
  scale_x_continuous(breaks = EG$x %>% unique %>% sort,
                     limits = c(0, 1.02)) +
  coord_cartesian(clip = 'off') +
  geom_text(
    data = EG %>% filter(N > 5),
    aes(x, y, label = N),
    size = 2.8,
    inherit.aes = FALSE,
    color = 'black'
  ) +
  scale_fill_distiller(palette = 'Blues',
                       direction = 1,
                       na.value = 'gainsboro')
```


In the end, we can observe the  proportion of tails fit per quadrant.
```{r, fig.width = 6, fig.height=5}
# Cuts based on median values
MC = median(results$coverage, na.rm = T)
MP = median(results$purity, na.rm = T)

# Classes for the plots
results$c_coverage = cut(
  results$coverage,
  breaks = c(-Inf, MC, Inf),
  labels = c("Low depth", "High depth")
)
results$c_purity = cut(
  results$purity,
  breaks = c(-Inf, MP, Inf),
  labels = c("Low purity", "High purity")
)

results$c_coverage = factor(results$c_coverage, levels = c("High depth", "Low depth"))

propo = results %>%
  group_by(c_coverage, c_purity) %>%
  summarise(p_tail = sum(pi_Tail > 0, na.rm = T) / n(),
            npt = 1 - p_tail) %>%
  arrange(p_tail) %>%
  ungroup()

propo = reshape2::melt(propo, id = c('c_coverage', 'c_purity'))

propo %>%
  group_by(c_coverage, c_purity) %>%
  mutate(lab.ypos = cumsum(value) - 0.5 * value) %>%
  ggplot() +
  geom_bar(
    aes(x = 1, y = value, fill = variable),
    stat = 'identity',
    width = 1,
    color = 'white'
  ) +
  facet_grid(c_coverage ~ c_purity) +
  mobster:::my_ggplot_theme() +
  ylim(0, 1) +
  labs(
    y = "Cases with tail (%)",
    x = '',
    caption = paste0("Median dept ", MC, ' - Median purity ', MP)
  ) +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(values = c(`p_tail` = 'indianred', `npt` = 'steelblue')) +
  coord_polar("y", start = 0) +
  geom_text(aes(
    y = lab.ypos,
    x = 1,
    label = paste0(round(value, 2) * 100, '%')
  ), color = "white") 
  
```