---
title: 'Simulated tumours: single-sample data'
author: "Giulio Caravagna and Timon Heide"
date: "12/03/2020"
output: 
  html_document:
    theme: cosmo
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE, warning=FALSE}
require(dplyr)
require(mobster)
```

This page describes the single-sample synthetic data used to test [MOBSTER](https://caravagn.github.io/mobster), where we fixed the simulated tumour sequencing (120x) and purity (100%). Some of these tumours are monoclonal neutral (0 subclones), others have 1 subclone, as discussed in the paper. 

The simulations have been collected in an [RDS object](./asymmetric_branching_simulations_batch1.1.rds), for conveniency. A [similar dataset](./asymetric_branching_simulations_batch1.1_variable_depth.rds) is released containing simulations with varying sequencing coverage.

**Note.** Data have been generated using the Tumour Emulator [TEMULATOR](https://t-heide.github.io/TEMULATOR) package developed by [Timon Heide](https://github.com/T-Heide) at the [Sottoriva Lab](http://www.sottorivalab.org/). See the simulator webpage to learn how to generate new sequencing data using a Markov model of stochastic tumour growth.

```{r}
simulations = readRDS("./asymmetric_branching_simulations_batch1.1.rds")
```

# Available cohort

```{r}
# Number of simulations
print(length(simulations))

# Simulation ids
print(names(simulations) %>% head)
```

# Analysis (example case)

We show how to fit one of the simulated cases, tumour `dcf11bece430ee3a98cc76ff0d91d2c67448258d`.

```{r, fig.width=6, fig.height=3}
id = 'dcf11bece430ee3a98cc76ff0d91d2c67448258d'

# Simulation data
tumour = simulations[[id]]

# The data structure contains different type of information
print(tumour)
```
Sequencing data  - see [TEMULATOR](https://t-heide.github.io/TEMULATOR) for details -  is in the format suitable for MOBSTER analysis. 

We here fit the model with the fast setup of MOBSTER (`auto_setup = 'FAST'`).

```{r, fig.width=6, fig.height=3, warning=F}
data = tumour$sequencing
head(data) %>% print

# fit by MOBSTER
fit = mobster_fit(data, auto_setup = 'FAST')

# Print and plot the fit
print(fit$best)
```

We can plot the fit.

```{r, fig.width=6, fig.height=4}
plot(fit$best)
```

The simulated tumour (its parameters, architecture etc) is stored in the object. In our tests, we consider detectable only clones that have >2.5% CCF (which corresponds to 5% VAF since we do not have copy number events); we also add explicitely the clonal cluster peak, because that is not recorded in the simulation of the initial clonal expansion.

```{r}
clone_peaks = round(tumour$cell_counts['fractions', 3:ncol(tumour$cell_counts)], 3)
clone_peaks = clone_peaks[clone_peaks > 0.05]

clone_peaks = c(clone_peaks, 0.5)

print(clone_peaks)
```

You can conclude that the subclone in this tumour has swept through almost completely (~45% VAF); the fits reports indeed one major clonal expansion (no evidence of ongoing subclonal selection, 1 Beta component).

### A polyclonal tumour (with a detectable subclone expanding)

This  example instead shows a clear sign of an ongoing subclonal selection.
```{r, fig.width=6, fig.height=4, message=FALSE, warning=FALSE}
id = "9583d8ac804182ed719509e00a1d1d07630a910f"

# fit by MOBSTER
fit = mobster_fit(simulations[[id]]$sequencing, auto_setup = 'FAST')
plot(fit$best)

```

The simulated CCF for the subclone is correctly identified
```{r}
round(
  simulations[[id]]$cell_counts[
    'fractions', 
    3:ncol(simulations[[id]]$cell_counts)
    ], 
  3) %>%
  print

print(fit$best)
```
