---
title: "GBM longitudinal analysis with MOBSTER"
author: "Giulio Caravagna"
date: "11/03/2020"
output: 
  html_document:
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
```


Table of fits for the glioblastoma (GBM) cohort published in:

* [Körber, et al. *"Evolutionary trajectories of IDHWT glioblastomas reveal a common path of early tumorigenesis instigated years ahead of initial diagnosis."* Cancer Cell 35.4 (2019): 692-704.](https://www.sciencedirect.com/science/article/abs/pii/S1535610819301023). 


Table files:

* [Summary analysis](Summary_GBM.tsv);

Tools and data used for this analyis:

* [MOBSTER](https://caravagn.github.io/mobster), a model-based method for tumour subclonal deconvolution;
* [BMix](https://caravagn.github.io/BMix), a Binomial mixture model for read counts clustering;
* Data provided by the authors of [PCAWG cohort](https://docs.icgc.org/pcawg/).

Note that:

* the tables contains the sample id (`case`), `type` refers to the two longitudinal samples available;
* `karyotype` identifies the segments copy state of the mutations analysed; the CNA calls  are provided by the authors;
* `K_beta` refers to information generated from  BMix analysis. 
* all other columns refer to MOBSTER. Columns named `Mean`/ `Variance` refer to parameters of the corresponding mixture components; columns named `N` to the number of mutations in the final output; columns `pi`  to the relative proportion of mutations and columns `K`  are the number of components fit for each statistical model (`K_B` and `K_BB` in BMix are Binomial/ Beta-Binomial mixtures, with `K_Beta` the Beta components fit by MOBSTER); tail information is reported in `tail` columns.



```{r, echo=FALSE, message=F,warning=FALSE}
library(DT)

datatable(
  read_csv('Summary_GBM.csv'),
  class = 'cell-border nowrap',
  filter = 'top',
  extensions = 'Buttons',
  options = list(
    pageLength = 12,
    AutoWidth = T,
    dom = 'Bfrtip',
    buttons = c('csv', 'excel', 'pdf', 'print'),
    scrollX = TRUE,
    fixedColumns = TRUE
  )
)
```
